Depletion capabilities in the OpenMC Monte Carlo particle transport code

### 2020-06-23 manuscript.tex
Add discussion of PWR results

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[charter]{mathdesign} % Type-1 Latin Modern modern
\usepackage[T1]{fontenc}         % Use T1 encoding instead of OT1
\usepackage[utf8]{inputenc}      % Use UTF8 input encoding
\usepackage{microtype}           % Improve typography
\usepackage{booktabs}            % Publication quality tables
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{float}
\usepackage{algorithm}
\usepackage{algorithmicx}
\usepackage{algpseudocode}
\usepackage{siunitx}
\usepackage{multirow}
\usepackage{upgreek}

@@

In addition to the full depletion chain, a simplified depletion chain based on
the VERA depletion benchmark specification~\cite{kim2015casl} was created and
made available for users at \url{https://openmc.org}. This chain includes 223
nuclides and has adjusted branching ratios for some nuclides to preserve
accuracy of nuclide production rates. \Cref{fig:chains} shows which nuclides are
included in each of the depletion chains.
\begin{figure}
  \centering
  \includegraphics[width=4in]{figures/chains.pdf}
  \caption{Table of nuclides showing which nuclides are included in each of the
    depletion chains based on ENDF/B-VII.1.}
  \label{fig:chains}
\end{figure}

@@

\subsubsection{Fission Product Yields}

@@

The abstract \texttt{DepSystemSolver} class represents a solver for
\cref{eq:depletion-matrix} when $\vect{A}$ is constant (a solver that can
evaluate \cref{eq:constant-A}). Subclasses of \texttt{DepSystemSolver} are
required to implement a method that takes the burnup matrix $\vect{A}$, an
initial density vector $\vect{n}_0$, and a timestep size, and returns the
resulting nuclide densities at the end of the step. The only subclass of
\texttt{DepSystemSolver} at present is \texttt{IPFCramSolver}, which implements
the IPF form of CRAM to evaluate \cref{eq:constant-A}. Because the CRAM method
requires solving multiple sparse linear systems, this class relies on the
third-party \texttt{scipy.sparse} package~\cite{virtanen2020nm} for performing a
sparse linear solve.

@@

@@

Using ENDF/B-VII.1 files from MCNP distribution. However, for light water
$S(\alpha,\beta)$, Serpent is not capable of handling continuous format, so the
file from ENDF/B-VII.0 was used. The evaluation didn't change between VII.0 and
VII.1, but the processing was different. For Serpent, decay and neutron fission
yield sublibraries are concatenated.

@@

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\usepackage[charter]{mathdesign} % Type-1 Latin Modern modern
\usepackage[T1]{fontenc}         % Use T1 encoding instead of OT1
\usepackage[utf8]{inputenc}      % Use UTF8 input encoding
\usepackage{microtype}           % Improve typography
\usepackage{booktabs}            % Publication quality tables
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{float}
\usepackage{subcaption}
\usepackage{algorithm}
\usepackage{algorithmicx}
\usepackage{algpseudocode}
\usepackage{siunitx}
\usepackage{multirow}
\usepackage{upgreek}

@@

In addition to the full and reduced depletion chains, a simplified depletion
chain based on the VERA depletion benchmark specification~\citep{kim2015casl}
was created and made available for users at \url{https://openmc.org}. This chain
includes 223 nuclides and has adjusted decay branching ratios for some nuclides
to preserve the accuracy of nuclide production rates. \Cref{fig:chains} shows
which nuclides are included in each of the depletion chains.
\begin{figure}
  \centering
  \includegraphics[width=4in]{figures/chains.pdf}
  \caption{Table of nuclides showing which nuclides are included in each of the
    depletion chains based on ENDF/B-VII.1.}
  \label{fig:chains}
\end{figure}

@@

\subsubsection{Fission Product Yields}
\label{sec:fpy}

@@

The abstract \texttt{DepSystemSolver} class represents a solver for
\cref{eq:depletion-matrix} when $\vect{A}$ is constant (a solver that can
evaluate \cref{eq:constant-A}). Subclasses of \texttt{DepSystemSolver} are
required to implement a method that takes the burnup matrix $\vect{A}$, an
initial density vector $\vect{n}_0$, and a timestep size, and returns the
resulting nuclide densities at the end of the step. The only subclass of
\texttt{DepSystemSolver} at present is \texttt{IPFCramSolver}, which implements
the IPF form of CRAM to evaluate \cref{eq:constant-A}. Because the CRAM method
requires solving multiple sparse linear systems, this class relies on the
third-party \texttt{scipy.sparse} package~\citep{virtanen2020nm} for performing
a sparse linear solve.

@@

To validate the depletion solver in OpenMC, we have carried out simulations
using the \texttt{openmc.deplete} module as well as Serpent 2.1.31 on two simple
problems. The first problem is a pressurized water reactor (PWR) pincell and the
second problem is a single assembly of a sodium-cooled fast reactor (SFR). These
problems were selected to cover both thermal and fast neutron spectra.

While comparison of simulation predictions to an experiment is a better means of
ensuring fitness for a particular application, it is not without downsides.
First, the presence of experimental uncertainties makes it hard to ascertain
whether small differences are attributable to the uncertainties or errors in the
code itself. In a cross-code comparison, if the same solution method and
fundamental nuclear data are being used in both codes, one expects to obtain the
same answer. Thus, even very minute differences may be indicative of a true
problem in the method implementation in one of the codes.

@@

Given the remarks above, it is of utmost importance to ensure that the same
nuclear data are being used in a cross-code comparison. For a depletion
simulation, we must ensure that the incident neutron cross sections and
secondary angle/energy distributions, radioactive decay data, and fission
product yields are the same for both codes. In the simulations performed, all
data was based on the ENDF/B-VII.1 nuclear data library~\citep{chadwick2011nds}.
Incident neutron cross sections from the ENDF/B-VII.1 library processed by the
LANL nuclear data team~\citep{conlin2013lanl}, known as endf71x, were used for
transport simulations in both OpenMC and Serpent. Serpent is able to use the ACE
files from the endf71x library directly while OpenMC uses an HDF5 format that
requires conversion of the ACE files first. For the PWR pincell problem,
$S(\alpha,\beta)$ data is also needed for hydrogen in light water; because
Serpent does not yet handle the continuous representation used in the ENDF71SaB
library, the light water data from ENDF/B-VII.0 was used instead. We note that
the light water evaluation didn't actually change between ENDF/B-VII.0 and
ENDF/B-VII.1.

In Serpent, the depletion solver relies directly on ENDF decay and FPY
sublibrary files. Appropriate decay and FPY libraries for Serpent were thus
formed by concatenating the ENDF files directly from the ENDF/B-VII.1
distribution. For OpenMC, cases were run with the full, reduced, and CASL
depletion chains based on ENDF/B-VII.1 decay and FPY sublibrary files.

To ensure that OpenMC and Serpent perform power normalization in the same
manner, the default fission heating values from Serpent were used in OpenMC. By
default, the energy release per fission in $^{235}U$ in Serpent is set to
\SI{202.27}{\mega\electronvolt}, and the values for other actinides is scaled
based on the Q value for fission listed in MF=3 in the ENDF evaluation. Without
the same heating values, the power normalization might result in OpenMC
depleting materials too quickly or too slowly.

Serpent uses energy-independent capture branching ratios by default that are
based on the JEFF 3.1 activation file using a PWR flux spectrum. For the PWR
problem, these energy-independent branching ratios were used in OpenMC. For the
SFR problem, Serpent was used to first generate representative capture branching
ratios in the fast flux spectrum using Serpent's energy-dependent capture
branching ratio treatment. Then, the energy-independent branching ratios were
set in both Serpent and OpenMC. As discussed in \cref{sec:fpy}, OpenMC also has
several methods for treating the energy dependence of FPYs. The OpenMC
simulations were carried out with the three different methods to see what effect
they have on resulting isotopic compositions as compared with Serpent.

@@

The PWR pincell problem is a single 2.4\% enriched PWR pincell from the BEAVRS
benchmark~\citep{horelik2013mc,horelik2018mit}. Material compositions and
dimensions from the BEAVRS benchmark were used without modification. The pincell
extends to positive/negative infinity in the $z$ direction and has reflective
boundary conditions in the $x$ and $y$ directions. A linear heating rate of
\SI{174}{\watt\per\centi\meter} was used for power normalization. The pin was
depleted to 50 MWd/kg using the following burnup schedule:
\begin{itemize}
  \item A single 0.1 MWd/kg, 0.4 MWd/kg, and 0.5 MWd/kg timesteps;
  \item 1.0 MWd/kg timesteps out to a cumulative burnup of 10.0 MWd/kg;
  \item 2.5 MWd/kg timesteps out to a cumulative burnup of 50.0 MWd/kg.
\end{itemize}
The short timesteps at the beginning are used to accurately capture the buildup
of xenon isotopes. In both codes, the predictor method was used for time
integration. While the predictor method is the least accurate time integration
method, the purpose of the comparison is to confirm that OpenMC and Serpent
produce comparable results, irrespective of the absolute values.

Six coupled transport-depletion simulations were carried out with OpenMC to test
a variety of combinations of depletion chains and methods for handling FPY
energy dependence:
\begin{table}

\end{table}
\begin{enumerate}
  \item Full depletion chain, thermal FPYs
  \item Full depletion chain, interpolated FPYs
  \item Reduced depletion chain, thermal FPYs
  \item CASL depletion chain, thermal FPYs
  \item CASL depletion chain, FPYs based on cutoff method
  \item CASL depletion chain, interpolated FPYs
\end{enumerate}
Because Serpent doesn't have an option for changing the depletion chain or the
method for handling FPY energy dependence, a single coupled transport-depletion
simulation with Serpent was carried out. For all simulations, 20 inactive
batches, 200 active batches, and \num{3.6e6} particles per batch were used,
which resulted in an uncertainty on the effective multiplication factor of about
3--4 pcm.

\Cref{fig:pwr-keff} shows the difference between $k_\text{eff}$ calculated by
OpenMC and Serpent. The blue points correspond to results using the full
depletion chain, the red points correspond to points using the reduced chain,
and the green points correspond to results using the CASL chain. For each chain,
the use of interpolated FPYs is denoted by circles and thermal FPYs by x's. At
zero days, the difference in $k_\text{eff}$ between OpenMC and Serpent is not
statistically significant. As the burnup increases, the difference for the full
chain cases rises to about 20 pcm. The reduced chain case exhibits the exact
same behavior as the full chain as one would expect. Use of the CASL chain
results in a slightly larger difference between OpenMC and Serpent at 50 MWd/kg,
about 40 pcm. We also see that for both the full and CASL chains, changing the
method for handling FPY energy dependence doesn't appear to have much of an
affect on $k_\text{eff}$, which is primarily driven by the prediction of
actinide concentrations.
\begin{figure}[H]
  \centering
  \includegraphics[width=4in]{figures/pwr_keff.pdf}
  \caption{Difference in the effective multiplication factor, $k_\text{eff}$,
  between OpenMC and Serpent for the PWR pincell model. While not shown, the one
  standard deviation uncertainty in the difference is about 3.5 pcm.}
  \label{fig:pwr-keff}
\end{figure}

\Cref{fig:pwr-actinides-full} shows the relative difference in predicted
actinide concentrations between OpenMC and Serpent at four different burnup
values. We see that most of the differences are less than 0.1\%, and all are
within 0.15\%. The differences in $^{235}$U and $^{239}$Pu concentrations
increase slightly with increasing burnup, which most likely accounts for the 20
pcm difference observed for the full chain in \cref{fig:pwr-keff}. While not
shown, the predicted actinide concentrations for the reduced chain are nearly
identical to those observed for the full chain in \cref{fig:pwr-actinides-full}.
\Cref{fig:pwr-actinides-casl} shows the differences in actinide concentrations
using the CASL depletion chain. Because the CASL depletion chain differs from
the full depletion chain primarily with respect to how fission products are
treated, the differences reported in \cref{fig:pwr-actinides-casl} closely
mirror those in \cref{fig:pwr-actinides-full}.
\begin{figure}[H]
  \centering
  \includegraphics[width=4in]{figures/pwr/pwr_actinides_full.pdf}
  \caption{Relative difference in actinide concetrations between OpenMC and
  Serpent for the PWR pincell model at 5, 12.5, 30, and 50 MWd/kg using the full
  depletion chain.}
  \label{fig:pwr-actinides-full}
\end{figure}
\begin{figure}[H]
  \centering
  \includegraphics[width=4in]{figures/pwr/pwr_actinides_casl.pdf}
  \caption{Relative difference in actinide concetrations between OpenMC and
  Serpent for the PWR pincell model at 5, 12.5, 30, and 50 MWd/kg using the CASL
  depletion chain.}
  \label{fig:pwr-actinides-casl}
\end{figure}

The predicted concentration of fission products depends on both the depletion
chain that is used and the method for handling the energy dependence of the FPY.
We begin with results for the full depletion chain and interpolated FPYs, which
most closely matches the method used in Serpent; \cref{fig:pwr-fp-full-average}
shows the relative difference in fission product concentrations between OpenMC
and Serpent for this case. Most of the differences are less than 0.2\%, and only
a few fission products show differences up to 0.5\%.
\begin{figure}[H]
  \centering
  \includegraphics[width=4in]{figures/pwr/pwr_fp_full_average.pdf}
  \caption{Relative difference in fission product concetrations between OpenMC
  and Serpent for the PWR pincell model at 5, 12.5, 30, and 50 MWd/kg using the
  full depletion chain and interpolated FPYs.}
  \label{fig:pwr-fp-full-average}
\end{figure}

\Cref{fig:pwr-fp-full-thermal} shows the predicted fission product concetrations
when using the full depletion chain along with thermal FPYs. In this case,
although most fission products are within 1\% of Serpent, they are
systematically overpredicted other than a few nuclides. Because Serpent uses the
average fission energy to interpolate between two sets of FPYs, the systematic
difference is actually expected when using thermal FPYs in OpenMC. As the
incident energy of a neutron causing fission increases, the ``saddle'' and
``wings'' in the FPY curve tend to increase significantly, and the peaks around
$A=95$ and $A=135$ tend to slightly decrease. This corresponds to exactly what
is observed in \cref{fig:pwr-fp-full-thermal}. Because thermal FPYs were used in
OpenMC for this case, we expect that fission products near one of the two peaks
of the FPY curve will be slightly overpredicted relative to Serpent, and fission
products away from the peak will be underpredicted. This can be confirmed for
particular nuclides. Take, for example, $^{109}$Ag; in the $A=109$ mass chain,
the nuclides with the highest independent yields are $^{109}$Tc and $^{109}$Mo.
Both these nuclides have independent yields that increase significantly with
increasing neutron energy. The same situation occurs for $^{129}$I, whose
predominant precursor is $^{129}$Sn. The independent yields for these three
precursor nuclides are shown in \cref{tab:fp-yields}.
\begin{figure}[H]
  \centering
  \includegraphics[width=4in]{figures/pwr/pwr_fp_full_thermal.pdf}
  \caption{Relative difference in fission product concetrations between OpenMC
  and Serpent for the PWR pincell model at 5, 12.5, 30, and 50 MWd/kg using the
  full depletion chain and thermal FPYs.}
  \label{fig:pwr-fp-full-thermal}
\end{figure}
\begin{table}[htbp]
  \centering
  \caption{Independent yields of select fission products due to fission in
  $^{235}$U (data from the ENDF/B-VII.1 FPY sublibrary).}
  \label{tab:fp-yields}
  \begin{tabular}{cccc}
    \toprule
    \textbf{Energy Spectrum} & \textbf{$^{109}$Tc} & \textbf{$^{109}$Mo} & \textbf{$^{129}$Sn} \\
    \midrule
    Thermal & 0.00013 & 0.00015 & 0.00230 \\
    Fast & 0.00034 & 0.00042 & 0.00550 \\
    Fusion & 0.00598 & 0.00109 & 0.01162 \\
    \bottomrule
  \end{tabular}
\end{table}

\Cref{fig:pwr-fp-casl-average} shows the relative difference in predicted
fission product concentrations between OpenMC and Serpent when using the CASL
depletion chain and interpolated FPYs. As in \cref{fig:pwr-fp-full-average},
almost all fission products are predicted to well within 1\% at 50 MWd/kg.
However, in this case there is one clear outlier: $^{85}$Kr. Upon further
investigation, it was determined that the root cause for this difference is the
fact that the CASL depletion chain uses a cumulative yield for $^{85}$Kr, and
the cumulative yield of $^{85}$Kr is known to be inconsistent with the
independent yields of the $A=85$ mass chain~\cite{pigni2015nds}. Serpent relies
solely on the independent yield in its depletion solver (as does the full
depletion chain). Therefore, there is nothing inherently wrong with the CASL
depletion chain or OpenMC; it is simply a matter of the underlying nuclear data
being inconsistent. The 5\% underprediction of $^{85}$Kr is also likely
responsible for the slightly higher difference in $k_\text{eff}$ for the CASL
depletion chain shown in \cref{fig:pwr-keff}.
\begin{figure}[H]
  \centering
  \includegraphics[width=4in]{figures/pwr/pwr_fp_casl_average.pdf}
  \caption{Relative difference in fission product concetrations between OpenMC
  and Serpent for the PWR pincell model at 5, 12.5, 30, and 50 MWd/kg using the
  CASL depletion chain and interpolated FPYs.}
  \label{fig:pwr-fp-casl-average}
\end{figure}

\Cref{fig:pwr-fp-casl-thermal} shows the relative difference in predicted in
fission product concentrations when using the CASL depletion chain along with
thermal FPYs. The results are quite close to those shown
\cref{fig:pwr-fp-full-thermal}, with the exception of $^{85}$Kr for the
aforementioned reasons.
\begin{figure}[H]
  \centering
  \includegraphics[width=4in]{figures/pwr/pwr_fp_casl_thermal.pdf}
  \caption{Relative difference in fission product concetrations between OpenMC
  and Serpent for the PWR pincell model at 5, 12.5, 30, and 50 MWd/kg using the
  CASL depletion chain and thermal FPYs.}
  \label{fig:pwr-fp-casl-thermal}
\end{figure}
